---
title: Ending the Mariners playoff drought.
author: Cody Zanga
date: '2022-10-28'
slug: mariners
categories: []
tags: []
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(ggalt)
library(gt)
#install.packages("mice")
library(mice)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
mariners <- read_csv("data/02_22_mariners.csv")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
SepMariners <- mariners %>% separate(`W-L`, into=c("Win", "Loss"), sep="-")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
MarinersWins <- SepMariners %>% mutate(
  Winner = as.numeric(Win),
  Loser = as.numeric(Loss),
)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
wins <- MarinersWins %>% 
  group_by (Year) %>% 
  summarise(TotalWins = sum(`W/L` == "W"),
            TotalRuns = sum(R),
            TotalRunsAgainst = sum(RA)) 
```

For those who don't follow sports or baseball closely, the Seattle Mariners held the longest playoff drought across all professional North American sports teams. Coming into 2022, the drought had reached 20 years having last made the playoffs during a historic 2001 season that saw the team win 116 games. 

Using data gathered from Baseball Reference, we're going to take a look at how the Mariners ended the drought in 2022. Along the way there's been changes to the playoff format as well as some extremely close calls that could have ended the playoff drought much earlier than when it finally did come to an end. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() +
  geom_step(data = wins, aes(x = Year, y = TotalWins)) +
  scale_y_continuous(breaks = c(30, 50, 70, 90)) +
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018, 2022)) +
  labs(title = "Mariners wins over the last 20 seasons.", 
       subtitle = "The Mariners broke their 20-year playoff drought in 2022.",
       x = "Year", y = "Total Wins", caption = "Source: Baseball-Reference | By Cody Zanga") +
   theme_minimal() + 
   theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 10), 
    plot.subtitle = element_text(size=12), 
    panel.grid.minor = element_blank()
    )
```

Going back to 2002, the year after the teams last playoff appearance, you can see that the the team won around or more than 90 games on a few occasions. Typically, in current day MLB if you get around 90 wins you're in a good position to make a playoff push. Before 2012, there was only one wild card team in each division, so less teams had a chance to get in. In 2002 and 2003, the Mariners won 93 games, but were kept out by a 99-win Anaheim team and a 95-win Boston team respectively.

In 2014 the Mariners won 87 games and were kept out on the last day of the season by an 88-win Oakland team. During the Covid shortened 2020 season that had an expanded playoffs, where only 60 games were played, the Mariners missed the playoffs once again by being two games behind Houston. Last year the Mariners won 90 games and fell short on the last two days of the season to Boston and New York, who both won 92 games.

```{r message=FALSE, warning=FALSE, echo=FALSE}
losses <- SepMariners %>% 
  group_by(Opp) %>%
  summarise(
    TotalLosses = sum(`W/L` == "L")) %>%
      arrange(desc(TotalLosses)) %>%
  top_n(10)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
mostwins <- SepMariners %>%
  group_by(Opp) %>%
  summarise(
    TotalWins = sum(`W/L` == "W")) %>%
      arrange(desc(TotalWins)) %>%
  top_n(10)

colnames(mostwins)[1] <- "Tm"
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
winslosses <- cbind(losses, mostwins)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
winslosses %>% gt() %>%
  cols_label(Opp = "Opponent", TotalWins = "Most Wins", Tm = "Opponent", TotalLosses = "Most Losses") %>%
  tab_header(title = "The Mariners beat up Oakland more than any other team over the last 20 years.",
             subtitle = "As you would expect, the divisional rivals are towards the top of these list.") %>%
  tab_style(style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")) %>%
  tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")) %>%
  tab_source_note(
    source_note = md("**By:** Cody Zanga  |  **Source:** [Baseball Reference](https://www.baseball-reference.com/teams/SEA/2022.shtml)")) %>%
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12))
   ) %>%
  opt_row_striping() %>% 
  opt_table_lines("none")
```

Here we can see the teams the Mariners have beat the most over the last 20 years as well as the teams that have won the most against them. The divisional rivals (Texas, Los Angeles, Oakland and Houston) are towards the top of both of these list and that makes sense because you play each of them 19 times over the course of the season, way more than all other opponents. You might notice Houston is way down on the most wins list and there's a couple reason for that. First off, the Mariners have had a tough time winning against them over the last handful of years and secondly, Houston didn't join the division until 2013.  

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() +
  geom_dumbbell(data = wins, aes(y=Year, x=TotalRuns, xend=TotalRunsAgainst),
                size = 2,
                colour = "grey",
                colour_x = "green",
                colour_xend = "red") +
  labs(title = "A bigger run differential usually means more wins for your team.", 
       subtitle = "2010 was an especially bad year for the Mariners",
       x = "Total runs scored vs. total runs against", y = "Year", caption = "Source: Baseball-Reference | By Cody Zanga") +
   theme_minimal() + 
   theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 10), 
    plot.subtitle = element_text(size=12), 
    panel.grid.minor = element_blank()
    )
```

Looking at this chart, we can see the amount of runs the Mariners have scored compared to the runs scored against them over this time period. We're essentially looking at the teams run differential, which is usually closely tied to how many games a team wins. One interesting fact is that team had a 90-win season in 2021 with a very poor run differential. The Mariners were able to create a positive run differential in 2022 and make the playoffs, just the first of hopefully many playoff runs for the team in the future.



